---
author: "Jake Sak"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE,
                      warning = FALSE, error = TRUE,
                      fig.height = 3)
library(tidyverse)
library(lubridate)
source("../../scripts/viridis.R")
```

## Assignment 6

#### Due Sunday, March 26, 11:59 PM CT

### Preliminaries

- Directories
    - COURSE/homework/
    - COURSE/homework/hw06/
    - COURSE/data/
    - COURSE/scripts/
- Files
  - COURSE/homework/hw06/hw06.Rmd
  - COURSE/data/exoplanets-clean-through-2022.csv
  - COURSE/data/wordle.csv
  - COURSE/scripts/viridis.R

### Aims

- Practice strings and regular expressions


### Problems

Problems 1--4 are based on a data frame with the names of built-in R colors.
  
```{r}
df_colors = tibble(color = colors()) %>% 
  mutate(index = row_number(), .before = color)
```

  1. Create a data frame with the names of all colors whose names are exactly 7 characters long. Print a summary of this data frame with the number of colors and the first and last alphabetically in the list.
  
```{r}
colors_7 = df_colors %>%
  filter(str_length(color) == 7)
colors_7 %>%
  summarize(num = n(), first = min(color), last = max(color)) %>%
  print()
```
  
  






  2. How many of the previous color names contain no digits? Create a second data frame which eliminates the rows from your solution to the previous problem any color names which contain a digit. Print a summary of this data frame with the number of colors and the first and last alphabetically in the list.

```{r}
colors_7_no_num = colors_7 %>%
  filter(!str_detect(color, "[:digit:]"))
colors_7_no_num %>%
  summarize(num = n(), first = min(color), last = max(color)) %>%
  print()
```





  3. Find all colors which contain consecutive letters "e". Print a summary with a count of how many such colors there are and the first and last alphabetically.
  
```{r}
colors_double_e = df_colors %>%
  filter(str_detect(color, "ee"))
colors_double_e %>%
  summarize(num = n(), first = min(color), last = max(color)) %>%
  print()
```



  
  
  4. Find all color names which contain any consecutive identical letters. Create a data frame with these colors.
  
- Print the first ten rows of this data frame
- Tabulate how often each letter is used in a consecutive letter pair and print out this table arranged in descending order of the counts.
- Identify the three color names which have a distinct pair of consecutive identical letters. (These are the only color names which use the given letter pair.)
  
```{r}
str_ident = df_colors %>%
  filter(str_detect(color, "([:alpha:])\\1"))
str_ident %>% print(n = 10)


str_ident_letters = str_ident %>%
  mutate(letter = str_extract_all(color, "([:alpha:])\\1")) %>%
  unnest(letter) %>%
  mutate(letter = str_sub(letter, 1, 1)) 

str_ident_letter_counts = str_ident_letters %>%
  group_by(letter) %>%
  summarize(count = n()) %>%
  arrange(desc(count)) %>% 
  print()


distinct_letters = str_ident_letter_counts %>%
  filter(count == 1) %>%
  select(letter)

distinct_letter_words = inner_join(str_ident_letters, distinct_letters, by = "letter")
```




  
  5. Load the exoplanet data set from the file `exoplanets-clean-through-2022.csv`. Create a new data frame by selecting only the columns `id` and `star`. Find and print all rows where the name in `star` does not match the string in `star` treated as a regular expression.
  
- What seems to be the common characteristic of such star names?

After identifying this characteristic, add a column to the `planets` data frame with the name `star_regex` which, when treated as a regular expression, does match the literal star name exactly. Such a regular expression should begin with `^` and end with `$`. Verify that each star name is matched by the regular expression.
  
  
  
```{r}
planets = read_csv("../../data/exoplanets-clean-through-2022.csv")
```
  

```{r}
no_match_stars = planets %>%
  select(id, star) %>%
  filter(!str_detect(star, star))
no_match_stars %>% print(n = Inf)

planets = planets %>%
  mutate(star_regex = str_c('^',star,"$")) %>%
  mutate(star_regex = str_replace_all(star_regex, "\\+", "\\\\+")) %>%
  mutate(star_regex = str_replace_all(star_regex, "\\.", "\\\\."))

planets %>% 
  summarize(not_matching = sum(!str_detect(star, star_regex)) )%>%
  print()

```

> All star names which do not match contain a + or a ., which have special characteristics in regular expressions




  6. Most planets have a name equal to the star name followed by a space and a single lower case letter. Identify all planets for which this pattern is not true.
  
- Print the columns `id`, `planet`, `star`, and `year` for all planets in `planets` which do not fit this pattern.
  
- How many of the 5240 exoplanets do not follow the typical convention?
  
  
```{r}
planets_no_star_name = planets %>%
  filter(!str_detect(planet, str_c(str_sub(star_regex, 1, -2), " ", "[:lower:]", "$"))) %>%
  select(id, planet, star, year) %>%
  print(n = Inf)
```

>48 of the 5240 exoplanets do not follow the typical convention
  
  



  7. For the planets whose names **do match** the pattern from the previous problem, create a table which counts the frequency of the last lower case letters in the planet name. Display this data with an effective graph.
  
  
  
```{r}
planet_letters = planets %>%
  filter(str_detect(planet, str_c(str_sub(star_regex, 1, -2), " ", "[:lower:]", "$"))) %>%
  mutate(planet_letter = str_sub(planet, -1, -1)) %>%
  group_by(planet_letter) %>%
  summarize(count = n())

ggplot(planet_letters, aes(x = planet_letter, y = count)) +
  geom_col(color = "black", fill = "blue") +
  scale_y_log10() +
  xlab("Planet Last Letter") +
  ylab("Count") +
  ggtitle("Last Letter Counts")
```




  8. The file `wordle_data.csv` contains the original solution list to the game Wordle.
In the game, players attempt to identify a secret five-letter word by guessing a word and using the color-coded information returned.
A green letter from the guess is in the correct location in the secret word.
A yellow letter from the guess is in the secret word, but is not in the correct location.
A gray letter is not in the word (unless a guess contains multiple copis of the same letter and at least one letter is colored and another is gray).

In a puzzle, the guess information is return for these first two guesses.

- SLATE returns the pattern Gray, Gray, Gray, Yellow, Gray
- COUNT returns the pattern Yellow, Green, Gray, Gray, Yellow

*(Note that the words in the data base have all lower case letters.)*

Find all possible solutions in `wrdl` which are consistent with the information given.

*Hint: you may find it easiest to pipe the word list through multiple `filter()` commands instead of trying to create a single regular expression which only matches all possible solutions. For example, you know the secret word contains at least one `T` and this letter is not in the 4th or 5th positions, at least one `C` which is not the first letter, and that the second letter is an `O`. You also know several letters which are not contained in the secret word.*

```{r}
wrdl = read_csv("../../data/wordle_data.csv") %>% 
  filter(solution == TRUE) %>% 
  select(word)
  
wrdl_possible = wrdl %>%
  filter(str_detect(word, "t")) %>%
  filter(str_detect(word, "c")) %>%
  filter(str_sub(word, 2, 2) == "o") %>%
  filter(!str_detect(str_sub(word, 4, 5), "t")) %>%
  filter(!str_starts(word, "c")) %>%
  filter(!str_detect(word, "s")) %>%
  filter(!str_detect(word, "l")) %>%
  filter(!str_detect(word, "a")) %>%
  filter(!str_detect(word, "e")) %>%
  filter(!str_detect(word, "u")) %>%
  filter(!str_detect(word, "n"))
wrdl_possible %>% print(n = Inf)
```




