---
author: "Jacob Sak"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE,
                      warning = FALSE, error = TRUE, fig.height = 3)
library(tidyverse)
library(kableExtra)
source("../../scripts/viridis.R")
source("../../scripts/ggprob.R")
```

\newcommand{\E}{\mathsf{E}}
\newcommand{\Var}{\mathsf{Var}}
\renewcommand{\prob}{\mathsf{P}}


## Assignment 7

#### Due Sunday, April 2, 11:59 PM CT

### Preliminaries

- Directories
    - COURSE/homework/
    - COURSE/homework/hw07/
    - COURSE/data/
    - COURSE/scripts/
- Files
  - COURSE/homework/hw07/hw07.Rmd
  - COURSE/scripts/viridis.R
  - COURSE/scripts/ggprob.R

### Aims

- Practice probability and the binomial distribution

### Problems

  1. A discrete random variable $X$ has possible values 0, 1, 2, 3, 4, 5, 6, 7, 8 with the following partial probability distribution.
The missing probabilities $\prob(X=7)$ and $\prob(X=8)$ are equal to each other.

```{r, echo = FALSE}
set.seed(20221021)
x = 0:8
p = rnorm(7,5,2.1)
p = sort(round(p / sum(p) * 0.84, 2))
p1 = tibble(x = x, p = c(p, rep(NA,2)))

p1_tab = p1 %>% 
  mutate(x = str_c(x)) %>% 
  pivot_wider(names_from = x, values_from = p) %>% 
  mutate(x = "P(X=x)", .before = `0`) 

p1_tab %>% 
  kable() %>% 
  kable_styling(position = "center", full_width = FALSE)
```

- Put the values `x = 0:8` and the tabled probabilities `p` into a data frame (use `tibble()`).

```{r}
full_p1 = tibble(
  x = 0:8,
  p = c(0.04, 0.06,	0.11,	0.12,	0.14,	0.17,	0.2, NA, NA)
)
```

- Write code to calculate the missing values for $\prob(X=7) = \prob(X = 8)$ and replace the missing values with these calculated values into your data frame.

```{r}
probs = full_p1 %>% select(p) %>% drop_na() %>% summarize(sum_p = sum(p)) %>% pull(sum_p)
remaining = 1 - probs
p_7and8 = remaining/2

full_p1 = full_p1 %>%
  mutate(p = case_when(
    !is.na(p) ~ p,
    TRUE ~ p_7and8))
full_p1 %>% print(n=Inf)
```

- Verify that the sum of all probabilities is 1.

```{r}
sum_probs = full_p1 %>%
  select(p) %>%
  summarize(sum(p)) %>% 
  print()
```

- Calculate the mean, variance, and standard deviation of this distribution.

```{r}
mean_var_p1 = full_p1 %>% 
  mutate(xp = x*p) %>% 
  summarize(x, p, mu = sum(xp)) %>%
  mutate(v = (x - mu)^2*p) %>%
  summarize(mu, var = sum(v), sd = sqrt(var)) %>%
  distinct() %>%
  print()
```






  2. Suppose you have a random variable $X \sim \text{Binomial}(120, 0.2)$.

- 2a. Calculate and report the mean and standard deviation of the distribution.

```{r}
n2 = 120
p2 = 0.2
mean_2 = n2*p2
sd_2 = sqrt(n2*p2*(1-p2))
print(mean_2)
print(sd_2)
```

- 2b. Calculate and report the probability that the random variable is exactly equal to 20, $\prob(X = 20)$.

```{r}
dbinom(20, n2, p2)
```

- 2c. Calculate and report the probability that $X$ equals 20 or more, $\prob(X \ge 20)$.

```{r}
prob_20_more = 1 - pbinom(19, n2, p2) 
prob_20_more %>% print()
pbinom(19, n2, p2, lower.tail = FALSE)
```

- 2d. Draw a graph of the binomial distribution with vertical lines at the mean (red and dashed), and 1, 2, and 3 standard deviations above and below the mean (red and dotted).


```{r}
gbinom(n2, p2, scale = TRUE) +
  geom_vline(xintercept = mean_2, color = "red", linetype = "dashed") +
  geom_vline(xintercept = mean_2 + c(-3, -2, -1, 1, 2, 3)*sd_2,
             color = "red", linetype = "dotted")
```





  3. What is the probability that $X$ from Problem 2 is within one, two, and three standard deviations of the mean? Round each probability to four decimal places.

```{r}
one_sd_prob = pbinom(mean_2 + sd_2, n2, p2) - pbinom(mean_2 - sd_2, n2, p2)
print(round(one_sd_prob, 4))
two_sd_prob = pbinom(mean_2 + sd_2*2, n2, p2) - pbinom(mean_2 - sd_2*2, n2, p2)
print(round(two_sd_prob, 4))
three_sd_prob = pbinom(mean_2 + sd_2*3, n2, p2) - pbinom(mean_2 - sd_2*3, n2, p2)
print(round(three_sd_prob, 4))

probs_q3 = pbinom(mean_2 + c(1,2,3)*sd_2, n2, p2) - pbinom(mean_2 - c(1,2,3)*sd_2, n2, p2)
print(round(probs_q3, 4))
```






  4. The following code makes a graph of the cumulative distribution function for the $\text{Binomial}(7, 0.3)$ distribution, that is $F(x) = \prob(X \le x)$. This is the function which is calculated by the base R function `pbinom()`.

- Modify the code to add a horizontal red dotted line at 0.4.


```{r}
prob4 = tibble(
  x = seq(-1,8,1),
  p = dbinom(x, 7, 0.3),
  cdf = pbinom(x, 7, 0.3))

ggplot(prob4, aes(x = x, y = cdf )) +
  geom_step(color = "blue") +
  geom_hline(yintercept = 0) +
  geom_hline(yintercept = 0.4, color = "red", linetype = "dotted") +
  scale_x_continuous(breaks = 0:7) +
  xlab("x") +
  ylab("Probability") +
  ggtitle("Binomial(7, 0.3) distribution CDF") +
  theme_minimal()
```

- Calculate the 0.4 quantile of the $\text{Binomial}(7, 0.3)$

```{r}
qbinom(0.4, 7, 0.3)
```

- Explain how the value of the quantile is related to the red dotted line which you added to the plot.

> The quantile is related to the red dotted line because the 0.4 quantile is 2, which is where the line at probability 0.4 intercepts the cdf

- Use the graph to determine the 0.75 quantile of this distribution.
Then verify your observation with a calculation.

> The 0.75 quantile according to the graph is 3

```{r}
qbinom(0.75, 7, 0.3)
```





  5. In a large population of likely voters, the proportion of voters who plan to vote for candidate A is $p = 0.52$. In a random sample of 600 likely voters, what is the probability that the proportion of voters who plan to vote for candidate A is less than 50% (the number of such voters is less than 300)?
  
```{r}
p5 = 0.52
n5 = 600
pbinom(299, n5, p5) %>% print()
```




  6. Basketball player 1 makes 80% of their free throws and basketball players 2 makes 70% of their free throws. Assume all of their free throws attempts are independent. If each player takes ten free throws and $X$ is the total number of free throws they make, does $X$ have a binomial distribution? If yes, specify the parameters $n$ and $p$. If not, explain why not.
  
> Yes, because it follows the BINS acronym, or has a fixed number of trials, distinct successes and failures, and independence. N for both players is equal to 10, or the number of attempts. P for player one is 0.8, and P for player 2 is 0.7, or the probability.




  7. What is the smallest number of coin tosses $n$ of a fair coin so that the chance of getting 50 or more heads exceeds 90%?  (You may need to do some trial and error. Alternatively, think of creating a data frame with a column `n` and a column for the probability and using this to find the answer.)

```{r}
heads_q7 = tibble(
 n = 0:1000,
 p = 0.5, 
 prob = 1 - pbinom(49, n, p)
) %>%
  filter(prob > 0.90) %>%
  select(n) %>%
  slice_min(order_by = n, n = 1) %>%
  pull(n)

heads_q7 %>% print()
```




  8. The random variable $X$ has the $\text{Binomial}(100, 0.2)$ distribution.
  
- Find an integer $a$ so that $P(X \le a) \ge 0.5$ and $P(X \ge a) \ge 0.5$.
- Show the values of $a$, $P(X \le a)$, and $P(X \ge a)$.

```{r}
a_val_q8 = tibble(
  n8 = 100,
  p8 = 0.2,
  a = 0:100,
  prob_less = pbinom(a, n8, p8),
  prob_more = 1 - pbinom(a-1, n8, p8)
) %>%
  filter(prob_less >= 0.5 & prob_more >= 0.5) %>%
  select(a, prob_less, prob_more) %>%
  rename("p(X<=a)" = prob_less, "p(X>=a)" = prob_more) %>%
  print()
```


