---
author: "Jacob Sak"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, fig.height = 4)
library(tidyverse)
library(lubridate)
library(scales)
library(modelr)
source("../../scripts/viridis.R")
source("../../scripts/ggprob.R")
```

\renewcommand{\prob}{\mathsf{P}}
\newcommand{\E}{\mathsf{E}}
\newcommand{\Var}{\mathsf{Var}}
\newcommand{\SD}{\mathsf{SD}}
\newcommand{\SE}{\mathsf{SE}}

## Homework Assignment 11

#### Due Friday, April 28, 2022, at 11:59 PM

### Preliminaries

- Directories
    - COURSE/homework/
    - COURSE/homework/hw11/
    - COURSE/data/
    - COURSE/scripts/
- Files
  - COURSE/homework/hw11/hw11.Rmd
  - COURSE/data/boston-marathon-data.csv
  - COURSE/data/dugong.csv
  - COURSE/scripts/viridis.R
  - COURSE/scripts/ggprob.R

### Data

- Some problems use the official Madison weather data, `madison-weather-official-1869-2022.csv`.
- Additional problems use the Boston Marathon data in the file `boston-marathon-data.csv`. This file is a transformed version of the raw data we used in class and has data for all runners who completed the race in 2010 and 2011. The variable `Time` is the sum of the times from the different portions of the race, each of which begins with "K".

### Aims

- Practice regression

## Problems

  **1.** In a regression problem to estimate $y$ from explanatory variable $x$ from a sample of size $n$, partial summary information is $\bar{x} = 20$ and $\bar{y} = 100$. Regardless of the values of other summary statistics, what is the value the predicted value $\hat{y}$ at a point where $x = 20$? Briefly explain.
  
> The prediction at x = 20 would be y = 100. This is because the regression line always passes through  $\bar{x}$,  $\bar{y}$.




  **2.** In a regression problem to estimate $y$ from explanatory variable $x$ from a sample of size $n$, partial summary information is $\bar{x} = 20$, $s_x = 5$, $\bar{y} = 100$, and $s_y = 15$. Which of the following values are possible values for the predicted value $\hat{y}$ when the explanatory variable has value $x = 30$? Briefly explain.
  
**(a)** 50      
**(b)** 70      
**(c)** 100      
**(d)** 120    
**(e)** 150

> The possible values for predicted value $\hat{y}$  when the explanatory variable has value $x = 30$ are (b) 70, (c) 100, and (d) 120. This is because we can standardize the formula for predicted values, where z = (30 - $\bar{x}$)/$s_x$. Then we plug this into the equation for $\hat{y}$, where $\hat{y}$ = $\bar{y}$ + r * z * $s_y$. because r has minimum -1 and maximum 1, we can plug these values in to find the bounds of $\hat{y}$.


Problems 3--6 are based on the data set in the file *dugong.csv* which relates age (in years) and length (in meters) of a sample of 27 dugongs, a type of marine mammal.
  
Credit:  The *dugong.csv* file is from Data8 at UC-Berkeley.


  **3.**

- Read in the *dugong.csv* data set.  
-  Create a scatter plot with `length` on the x-axis and `age` on the y-axis; be sure to add descriptive axis labels (include units of measurement) and a title.  
-  Using `geom_smooth()`, add the least-squares line to your plot.

```{r}
dugong = read_csv("../../data/dugong.csv")

ggplot(data = dugong, aes(x = Length, y = Age)) +
  geom_point() +
  geom_smooth(se = FALSE, method = "lm") +
  xlab("Length (meters)") +
  ylab("Age (years)") +
  ggtitle("Age Vs. Length of Dugongs")

```





  **4.**

- Using the dugong data, calculate the sample means, sample standard deviations, and correlation coefficient of the variables `age` and `length`.
- Using formulas from lecture, calculate the slope and intercept of the least squares regressions line to predict age with length.

```{r}
dugong_stats = dugong %>%
  summarize(xbar = mean(Length),
            sdx = sd(Length),
            ybar = mean(Age),
            sdy = sd(Age),
            r = cor(Length, Age)) %>%
  print()
slope = dugong_stats$r * (dugong_stats$sdy/dugong_stats$sdx)
intercept = dugong_stats$ybar - slope*dugong_stats$xbar
slope
intercept
```

- Use the dugong data and the functions `lm()` and `coef()` to calculate the slope and intercept of the least squares regression line of age against length (use length to predict age).

```{r}
dugong_lm = lm(Age~Length, data = dugong)
cf = coef(dugong_lm)
cf
```

- Verify that you get the same values.






  **5.**

- Add columns with the predicted values and residuals to the dugong data set. *(You can use* **modelr** *functions or just use `mutate()` and calculate these values directly.)*
- Plot the residuals versus length.
    - Add a horizontal line at $y=0$ and appropriate labels on each axis.

```{r}
library(modelr)
dugong = dugong %>%
  add_residuals(dugong_lm) %>%
  add_predictions(dugong_lm)

ggplot(dugong, aes(x=Length, y =resid)) +
  geom_point() +
  xlab("Length") +
  ylab("Residuals") +
  ggtitle("Residuals Vs. Length") +
  geom_hline(aes(yintercept=0), color="red", linetype = "dashed")
```

- Describe what the residual plot suggests about the appropriateness of using simple linear regression to predict age from length of dugongs.

> The residual plot looks to have a mean at around 0 and a fairly constant standard deviation. there are a few outliers that skew thew data due to a relatively small sample size, but it appears that the linear model is pretty accurate for the data.






  **6.**

- Print the summary of the fitted regression model

```{r}
summary(dugong_lm)
```

- The simple linear regression model for $Y_i$ conditional on the values of $X_i = x_i$ is

$$
\E(Y_i \mid X_i = x_i) = \beta_0 + \beta_1 x_i + \varepsilon_i, \quad \text{for $i = 1, \ldots,n$}
$$

where $\varepsilon_i \sim \text{Normal}(0, \sigma)$
for some parameter $\sigma > 0$.

- The parameter $\sigma$ is the unknown population standard deviation of the typical distance between a point $Y_i$ and its true expected value.

- Use the function `sigma()` on the fitted regression object (what you created with `lm()`) to extract the estimate of $\sigma$. Identify where this numerical value appears in the printed summary you made earlier.

```{r}
sigma(dugong_lm)
```
> This value is equal to the residual standard error

- The numerical estimate of $\sigma$ here is not quite the standard deviation of the residuals because the denominator is $n-2$, the degrees of freedom in simple linear regression, instead of $n-1$, the degrees of freedom from a single numerical sample.

- Use the column of residuals in the augments data set `dugong` and verify that:
    - the mean of the residuals equals zero (numerically, it might be very close).
    - you arrive at the numerical estimate of $\sigma$ by calculating
    
$$
\sqrt{ \frac{\sum_{i=1}^n (y_i - \hat{y}_i)^2}{n-2} }
$$

where the $i$th residual is $y_i - \hat{y}_i$.

```{r}
mean(dugong$resid)
sqrt(sum(dugong$resid^2)/(length(dugong$resid) - 2))
```







- Problems 7--8 use the cleaned Boston Marathon data in `boston-marathon-data.csv`.


  **7.**

- Read in the Boston marathon data from the file `boston-marathon-data.csv`.

```{r}
bm = read_csv("../../data/boston-marathon-data.csv")
```

- Create a scatter plots of `Time` versus `Age` for the female runners in 2010.
    - Add a straight regression line
    - Add a smooth curve
- As there are so many points, you may set `alpha` to a value less than one inside of `geom_point()` to lessen the effects of overplotting.    
    
```{r}
bm_q7 = bm %>%
  filter(Sex == "female", Year == "2010")

ggplot(data = bm_q7, aes(x = Age, y = Time)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se = FALSE) +
  geom_smooth(se = FALSE, method = "lm", color = "magenta") +
  xlab("Age (years)") +
  ylab("Finishing time (minutes)") +
  ggtitle("Boston Marathon finishing times of females by age 2010")
```
    
- Make a residual plot of the residuals versus `Age`.
    - Include a horizontal line at $y=0$
    - Include a smooth curve through the residuals

- In addition, make a density plot of the residuals.    
```{r}
bmq7_lm = lm(Time~Age, data = bm_q7)
bm_q7 = bm_q7 %>%
  add_residuals(bmq7_lm) %>%
  add_predictions(bmq7_lm) %>%
  select(Time, Age, resid, pred)

ggplot(bm_q7, aes(x=Age, y =resid)) +
  geom_point() +
  xlab("Age") +
  ylab("Residuals") +
  ggtitle("Residuals Vs. Age") +
  geom_hline(aes(yintercept=0), color="red", linetype = "dashed") +
  geom_smooth(se = FALSE)

ggplot(bm_q7, aes(x=resid)) +
  geom_density() +
  xlab("Residuals") +
  ylab("Density")
```







  **8.** Examine the residual plots from the previous problem.
  
- Is there evidence of strong non-linearity?

> As shown in the smooth curve through the residuals, there is some non-linearity, but it looks overall fairly linear

- Is there evidence that the standard deviation of the residuals varies substantially with changes in age?

> The standard deviation seems to shrink a little as age gets older, but I wouldnt say there is substantial change of the standard deviations


- Is there evidence that the error distribution for individual residuals is not symmetric?

> There is a distinct right skew on the residual density plot that would appear to prove the residual data is not symmetric


