---
title: "Project Draft"
author: "Jacob Sak, Ayush Kumar, Minji Suh, Paige Ellingson"
date: "2023-04-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE,
                      warning = FALSE, error = TRUE, fig.height = 3)
library(tidyverse)
library(lubridate)
library(kableExtra)
library(broman)
source("../../scripts/viridis.R")
source("../../scripts/ggprob.R")
theme_set(theme_minimal())
```

```{r include=FALSE}
meteorites <- read_csv("meteorite_landings.csv")
```

```{r include=FALSE}
head(meteorites, 10)
```


## Introduction
Back in 2015, the Meteoritical Society released their first iteration of the meteorite landing dataset, which includes the name, class, mass, year of the fall, and location (in longitude and latitude). The society continued to update their dataset with the most recent updates being uploaded on January 31, 2023.

While the dataset includes a lot of information, we find that the data lacks analysis into the comparison between certain columns. The group's motivation behind analyzing is to investigate the data and answer the question, "Is there a difference in mean mass between the meteorites that fell relative to the meteorites that were found by class?"


## Background
Although this dataset was created back in 2015, the Meteoritical Society, along with other organizations, have been collecting meteorite data ranging from 861 (earlier meteorite landings are collected via historical sources or carbon dating) to the present. Today, meteorites are observed in 2 ways:
1. Fell: Meteorite is first observed at some point while falling
2. Found: Meteorite is not observed falling but is found at some point after landing

Each meteorite in the dataset furthermore has a defined class, mass, location of landing (in terms of longitude and latitude), and fall status. 

Here is the source of the meteorite landing data: [Meteorite Landings](https://data.nasa.gov/Space-Science/Meteorite-Landings/gh4g-9sfh)

For the rest of this report, we will provide a means inference of the meteorite dataset as well as a distribution and a confidence interval.


## Analysis

#### Summary Statistics
```{r}
#Data Analysis
meteorites_clean = meteorites %>%
  drop_na(`mass (g)`) %>%
  drop_na(fall) %>%
  mutate(`mass (kg)` = `mass (g)` / 1000) %>%
  select(name, id, `mass (kg)`, fall)

meteor_fall_sum = meteorites_clean %>%
  filter(fall == "Fell") %>%
  summarize(xbar = mean(`mass (kg)`), sd = sd(`mass (kg)`), n = n())
print(meteor_fall_sum)
meteor_found_sum = meteorites_clean %>%
  filter(fall == "Found") %>%
  summarize(xbar = mean(`mass (kg)`), sd = sd(`mass (kg)`), n = n())
print(meteor_found_sum)

```

#### Visualization of the distribution

```{r}
#Data distribution
ggplot(meteorites_clean, aes(y = fall, x = `mass (kg)`, fill = fall)) +
  geom_boxplot() +
  ylab("Method") +
  xlab("Mass (kg)") +
  ggtitle("Meteor Mass Spread by Observation Method") +
  guides(fill=guide_legend(title="Method")) +
  scale_x_log10(n.breaks = 10) +
  theme_minimal()
```

```{r}
ggplot(meteorites_clean, aes(x = `mass (kg)`, fill = fall)) +
  geom_density(alpha = 0.75) +
  xlab("Mass (kg)") +
  ylab("Density") +
  ggtitle("Fell vs. Found Meteor mass distributions") +
  guides(fill=guide_legend(title="Method")) +
  scale_x_log10(n.breaks = 10) +
  theme_minimal()
```

## Confidence Interval for the Difference in Means and Inference on Difference in Means

```{r}
#t.test
meteor_fall_data = meteorites_clean %>%
  filter(fall == "Fell") %>%
  pull(`mass (kg)`)
meteor_found_data = meteorites_clean %>%
  filter(fall == "Found") %>%
  pull(`mass (kg)`)
meteor_ttest = t.test(meteor_fall_data, meteor_found_data)
print(meteor_ttest)
```

#### Confidence Interval
```{r}
#Method 1
ci1 = meteor_ttest$conf.int
print(ci1)

#Method 2
dof = t.test(meteor_fall_data, meteor_found_data)$parameter
ci2 = meteor_fall_sum$xbar - meteor_found_sum$xbar + c(-1,1)*qt(0.975, dof)*sqrt(meteor_fall_sum$sd^2/meteor_fall_sum$n + meteor_found_sum$sd^2/meteor_found_sum$n)
print(ci2)
```
>We are 95% confident the difference in mean mass of meteors that fell versus those that were found is between -8.63066 and 77.84825 (fell - found).


#### Inference on Difference in Means

**Statistical model:**

  - $F$ is an unspecified distribution, 
  - $\mu_1$ and $\sigma_1$ are the mean and standard deviation of the meteors that fell
  - $\mu_2$ and $\sigma_2$ are the mean and standard deviation of the meteors that were found

$$
X_i \sim F(\mu_1, \sigma_1), \ \ \ i = 1, 2, ..., n_1 \\
Y_j \sim F(\mu_2, \sigma_2), \ \ \ j = 1, 2, ..., n_2
$$

**Test Statistic:**

$$
T = \frac{\bar{x} - \bar{y}}{\sqrt{\frac{s_1^2}{n_1}+\frac{s_2^2}{n_2}}}
$$

**Hypotheses:**
$$
H_0: \mu_1 - \mu_2 = 0 \\
H_a: \mu_1 - \mu_2 \ne 0
$$

**Testing:**

```{r}
#Means inference

#Method 1
p_value1 = meteor_ttest$p.value
print(p_value1)

#Method 2
se = sqrt(meteor_fall_sum$sd^2/meteor_fall_sum$n + meteor_found_sum$sd^2/meteor_found_sum$n)
tstat = (meteor_fall_sum$xbar - meteor_found_sum$xbar) / se
pvalue2 = 2*pt(-abs(tstat), dof)
print(pvalue2)
```
**Interpretation of Results:**

With 95% confidence, there is insufficient evidence to reject the null hypothesis that the mean masses of meteors observed when falling and when found are different. In other words, we cannot conclude that there is a significant difference in the mean masses (p = 0.1165918, df = 1107.2).


## Discussion
Our analysis shows that there is no difference between the mean mass of meteorites that fell and the mean mass of meteorites that were found in the long run. This result further implies that over time, the average mass of any meteorites that has were observed when falling onto the Earth vs first observed when found on the Earth should be similar. 

Because our data does not cover all meteorites in space, it is possible we havev a sampling error and our sample means are not a true representation of all meteors. Moreover, the data of mass only includes meteorites that fell and were found in a particular time range. This might also fail to interpret the future mass of meteorites due to an error in population sampling.

Additional work needs to be made to successfully imply future meteorites that will be landing and found on the Earth. Successful assumptions will protect people on Earth and prepare for landings based on the size of the meteorite. To address this work, the population should be expanded to the space from Earth. The new question could be: "What is the difference in mean mass between meteorites on Earth and space? The result of this question could interpret any future meteorite that will be landing or found on the Earth or destroyed in space. 

In conclusion, there is no difference in mean mass between the meteorites that fell relative to the meteorites that were found. Our p-value (p = 0.1165918) proves that there is no significant evidence to reject the null hypothesis that the mean masses of meteorites that have been fallen and found are different.


## References